Rationality, Evolution \& the Environment

### 2015-08-21 paper/paper.tex
results section edits

\subsection{What should be in this section?}

\myalert{main purpose of this section: introduce max EU and min regret with some background,
  definitions, simple example and reasons why we care to look at especially these}

\begin{itemize}
  \item \todo[inline]{choose uniform consensual terminology for maximization of EU and regret minimization}
\item classical notion of rationality: maximization of expected utilities
  \begin{itemize}
  \item human choice behavior seems to deviate from classical rational choice in many ways
    \citep[most
    notably][]{TverskyKahnemann1974:Judgement-under,KahnemannTversky1979:Prospect-Theory}
  \end{itemize}
\item historical competitor: regret minimization
  \begin{itemize}
  \item introduced by Savage \& Niehans independently
  \item worked out (in parameterized form) by \citet{LoomesSugden1982:Regret-Theory:-} and
    argued to outperform prospect theory
  \item recently rediscovered by game theorists \citep{HalpernPass2012:Iterated-Regret} as a
    possible partial explanation of some puzzles (e.g., Travellers Dilemma)
  \end{itemize}
\item definitions of max EU and min reg (with minimal formal notation); simple example where
  they differ
\item motivate why we are interested in regret minimization; e.g.:
  \begin{itemize}
  \item simple historical alternative; certain prima facie plausibility
  \item argued to be empirically adequate (Loomes \& Sugden, Bell)
  \item gives us intriguing results that nicely motivate the general approach of meta-games
  \item non-independence of choices also for non-human decision makers (cite)
  \end{itemize}
\item here we care about game problems, not solitary decision problems
\end{itemize}

\subsection{Version 1}

@@

DEFINITION(SEU-maximization): Let $S$ be a set of states of the world, $X$ a set of outcomes and $A$ a set of actions. Given a probability measure $P$ on $S$ and a subjective utility function $u:X \rightarrow \mathbb{R} $, an action $a^*: S \rightarrow X $ is SEU-maximizing if $a^* \in \text{argmax}_{a_i \in A} \int_S u(a^*(s))dP(s) $. \\

@@

Results from Section~\ref{sec:results:-basic-model} relied heavily on
Fact~\ref{fact:equivalence2x2} that, for the special case of $2 \times 2$ symmetric games,
regret minimization $\tuple{\tau^{reg}, \Delta(A)}$ is behaviorally equivalent with expected
regret minimization $\tuple{\tau^{reg}, \bar{\mu}}$ and expected utility maximization
$\tuple{\tau^{\pi}, \bar{\mu}}$. This is no longer true when we look at arbitrary $n \times n$
games with $n \ge 2$. Table~\ref{tab:ExpectedFitness_10x10} therefore gives approximations for
expected fitness in a meta-game over a class of $n \times n$ symmetric games where
$n$ is randomly drawn from $\set{2, \dots, 10}$. Concretely, the numbers in
Table~\ref{tab:ExpectedFitness_10x10} are averages of payoffs obtained in 100,000 randomly
sampled games, where each game was sampled by first picking a number of acts
$n \in \set{2, \dots, 10}$ uniformly at random, and then filling the resulting $n \times n$
payoff matrix of the stage game with i.i.d.~sampled payoffs as before.

@@

>>>>

@@

\begin{definition}[SEU-maximization] Let $S$ be a set of states of the world, $X$ a set of
  outcomes and $A$ a set of actions. Given a probability measure $P$ on $S$ and a subjective
  utility function $u:X \rightarrow \mathbb{R} $, an action $a^*: S \rightarrow X $ is
  SEU-maximizing if $a^* \in \text{argmax}_{a_i \in A} \int_S u(a^*(s))dP(s) $.
\end{definition}

@@

Results from Section~\ref{sec:results:-basic-model} relied heavily on
Fact~\ref{fact:equivalence2x2} that, for the special case of $2 \times 2$ symmetric games,
regret minimization $\tuple{\tau^{reg}, \Delta(A)}$ is behaviorally equivalent with expected
regret minimization $\tuple{\tau^{reg}, \bar{\mu}}$ and expected utility maximization
$\tuple{\tau^{\pi}, \bar{\mu}}$. This is no longer true when we look at arbitrary $n \times n$
games with $n \ge 2$. We should therefore see what happens in a broader class of games.

Table~\ref{tab:ExpectedFitness_10x10} therefore gives approximations for expected fitness in a
meta-game over a class of $n \times n$ symmetric games where $n$ is randomly drawn from
$\set{2, \dots, 10}$. Concretely, the numbers in Table~\ref{tab:ExpectedFitness_10x10} are
averages of payoffs obtained in 100,000 randomly sampled games, where each game was sampled by
first picking a number of acts $n \in \set{2, \dots, 10}$ uniformly at random, and then filling
the resulting $n \times n$ payoff matrix of the stage game with i.i.d.~sampled payoffs as
before.

@@

As is to be expected, average payoffs obtained in games with more than 2 acts are higher, as
can be seen by comparing Table~\ref{tab:ExpectedFitness_10x10} with
Table~\ref{tab:ExpectedFitness_2x2_Full}. But increases in approximate expected fitness are not
uniform. Most importantly, the regret minimizing type $\tuple{\tau^{reg}, \Delta(A)}$ is
strictly dominated by $\tuple{\tau^{reg}, \bar{\mu}}$ and by $\tuple{\tau^{\pi}, \bar{\mu}}$ in
the meta-game from Table~\ref{tab:ExpectedFitness_10x10}. This means that while regret
minimization \emph{can} thrive in some evolutionary contexts, there are also contexts where it
is demonstrably worse off.

Still, although this may be bad news for regret minimizing types
$\tuple{\tau^{reg}, \Delta(A)}$, it is not the case that regret types \emph{as such} are weeded
out by selection. Since, by Fact~\ref{fact:maxEU-minReg}, $\tuple{\tau^{reg}, \bar{\mu}}$ and
$\tuple{\tau^{\pi}, \bar{\mu}}$ are generally behaviorally equivalent, it remains that
selection based on meta-games constructed from $n \times n$ games will still not eradicate all
regret types. So, although regret types may not be selected for in this case, they are also not
selected against.

On the other hand, there are plenty of ways in which the basic insight from
Proposition~\ref{proposition1} that regret-based preferences can be strictly better than
veridical objective preferences in case of radical uncertainty could make for situations in
which evolution would select for regret types exclusively, even in $n \times n$ games. If, for
example, epistemic types of players are not what biological evolution selects, but rather
something that the particular choice situation would exogenously give us, then regret-based
preference \emph{can} again drive out veridical preferences altogether. For concreteness,
suppose that only preference types compete and that agents' epistemic types are exogenously
given, in such a way that agents hold flat probabilistic beliefs $\bar{\mu}$ with probability
$1-p$ and are of type $\Delta(A)$ with probability $p$. This transforms the meta-game from
Table~\ref{tab:ExpectedFitness_10x10} into the simpler $4 \times 4$ game in which the payoff
obtained for a preference type is the weighted average over the payoffs of that preference
types in Table~\ref{tab:ExpectedFitness_10x10}. Setting $p = 0.02$ for illustration, we get the
meta-game in Table~\ref{tab:ExogeneousEpistemics}.

\begin{table}[]
\centering
\begin{tabular}{ccccc}
  \toprule
  & $\tau^{reg}$ 
  & $\tau^{\pi}$ 
  & $\tau^{com}$
  & $\tau^{alt}$ \\ 
  \midrule
  $\tau^{reg} $ & 6.926 & 6.926 & 5.942 & 7.757 \\ 
  $\tau^{\pi} $ & 6.924 & 6.924 & 5.948 & 7.751 \\ 
  $\tau^{com }$ & 6.566 & 6.570 & 5.481 & 7.434 \\ 
  $\tau^{alt} $ & 6.463 & 6.461 & 5.478 & 7.469 \\ 
   \bottomrule
\end{tabular}
\caption{Meta-game for the evolutionary competition between preference types when epistemic types are exogenously
  given}
\label{tab:ExogeneousEpistemics}
\end{table}

The only evolutionarily stable state of this meta-game is the regret type. Accordingly, all of
our simulation runs of the (discrete-time) replicator dynamic converged to monomorphic
regret-type populations. The reason why regret types prosper is because they have a substantial
payoff advantage if paired with radical uncertainty. If radical uncertainty is exogenously
given as something unavoidable that happens to agents (e.g., whenever they really have no
indication as to what the situation holds for them), and even if that happens only very
infrequently (i.e., for rather low $p$), regret types \emph{can} outperform veridical
preference types, as well as competitive and altruistic types.
